---
title: "Programming.cleaned.up"
author: "Elias Joneborg"
date: "3/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}

#Necessary packages
library(dplyr)
library(plyr)
library(incidence)
library(scales)
library(ggplot2)
library(reshape2)
library(ggthemes)
library(DiagrammeR)
library(table1)
library(jtools)
library(kableExtra)
library(ggstance)


swetrau_scrambled <- read.csv("swetrau-scrambled.csv")
problem_scrambled <- read.csv("problem-scrambled.csv")

problem_scrambled$OFI <- with(problem_scrambled, ifelse(Problemomrade_.FMP == "ok" | Problemomrade_.FMP == "OK" | Problemomrade_.FMP == "Ok" | Problemomrade_.FMP == "Föredömligt handlagd", "NO", "YES"))
#renaming and changing output of Problemområde.FMP to OFI and YES/NO

swetrau_problem_merged<- merge(swetrau_scrambled, problem_scrambled, by= "id", all.y = TRUE)
#merge databases

known_problem_area <-swetrau_problem_merged[!is.na(swetrau_problem_merged$OFI),]
#remove uknown problem area

fewer_variables<-known_problem_area %>% select(id,inj_dominant, pt_age_yrs, OFI, NISS, ed_gcs_sum, pre_intub_type, pre_sbp_value, pre_sbp_rtscat, ed_sbp_value, ed_sbp_rtscat, Ankomst_te, starts_with("AISCode"))
#removing unneccessary variables


## Changing 999 to NA for grouping into cohorts

fewer_variables$ed_sbp_value[fewer_variables$ed_sbp_value == 999] = NA
fewer_variables$ed_sbp_rtscat[fewer_variables$ed_sbp_rtscat == 999] = NA
fewer_variables$ed_gcs_sum[fewer_variables$ed_gcs_sum == 999] = NA
fewer_variables$inj_dominant[fewer_variables$inj_dominant == 999] = NA

### COHORT GROUPING ###

## The functions below expect a row of AIS codes as input

#BLUNT MULTISYSTEM
is_serious <- function(code) {
  severity <- substr(as.character(code), 8, 8)
  as.numeric(severity) >= 3  
}
#if severity level is 3 or higher, the injury is considered serious

number_of_regions <- function(code) {
  region <- substr(as.character(code), 1, 1)
  length(unique(region))
}
#identifying number of regions

has_more_than_one_serious_injury <- function(code) {
  code <- code[!is.na(code)]
  serious_injury <- is_serious(code)
  number_of_regions(code[serious_injury]) >= 2
}

## combining them
blunt.multisystem <- apply(fewer_variables[1:100, ], 1, function(row) {
    code <- row[grep("AISCode", names(row))]
    row
})

fewer_variables$blunt_multisystem <- NA
#creating column blunt_multisystem

known_dominant_injury <-fewer_variables[!is.na(fewer_variables$inj_dominant), ] 
#removing not known in inj_dominant

for (i in 1:nrow(known_dominant_injury)) {
  v = 0+i
  if (has_more_than_one_serious_injury(known_dominant_injury[v,grepl( "AISCode" , names(known_dominant_injury))]) == TRUE && known_dominant_injury[v,"inj_dominant"] == 1) {
    known_dominant_injury[v,"blunt_multisystem"] <- TRUE
  } else {
    known_dominant_injury[v,"blunt_multisystem"] <- FALSE
  }
}
#blunt multisystem <- TRUE/FALSE 

## The above for loop could be replaced with the more R:ish apply function below, which is substantially faster (at least on my low performing machine)

known_dominant_injury$blunt_multisystem <- apply(known_dominant_injury, 1, function(row) {
    code <- row[grep("AISCode", names(row))] 
    has_more_than_one_serious_injury(code) && row["inj_dominant"] == "1"
})

#SHOCK
bp_is_na <- with(fewer_variables, is.na(ed_sbp_rtscat) & is.na(ed_sbp_value))    
known_blood_pressure <- fewer_variables[!bp_is_na,]  
#removing rows with na in both sbp value and rts
     
known_blood_pressure$ed_sbp_rtscat[is.na(known_blood_pressure$ed_sbp_rtscat)]<-10
known_blood_pressure$ed_sbp_value[is.na(known_blood_pressure$ed_sbp_value)]<-999
#an ineffective way to ensure I don't get na:s in "ed_sbp_below_90" column below

known_blood_pressure$ed_sbp_below_90 <- with(known_blood_pressure, ifelse(ed_sbp_value>=0 & known_blood_pressure$ed_sbp_value<=90, TRUE, FALSE))
known_blood_pressure$shock <- with(known_blood_pressure, ifelse(ed_sbp_below_90 == TRUE | ed_sbp_rtscat <= 3, TRUE, FALSE))
#creating column for bp<90 with TRUE/FALSE

#GERIATRIC
known_age <- fewer_variables
known_age$geriatric <- with(fewer_variables, ifelse(pt_age_yrs>65 & pt_age_yrs<120, TRUE, FALSE))
#creating column in new df with TRUE/FALSE

#PENETRATING
neck_chest_abdomen_region <- function(code) {
  code <- code[!is.na(code)]
  region <- substr(as.character(code), 1, 1)  
  is.element(region,3:5)
}
#generates TRUE/FALSE

serious_neck_chest_abdomen_injury <- function(code) {
  code <- code[!is.na(code)]
  serious_injury <- is_serious(code)
  sum(neck_chest_abdomen_region(code[serious_injury]) == TRUE )
}
#generates numerical value

known_dominant_injury_pen <-fewer_variables[!is.na(fewer_variables$inj_dominant), ] 
known_dominant_injury_pen$penetrating <- NA
#removing na in dominant injury, new df

for (i in 1:nrow(known_dominant_injury_pen)) {
  v = 0+i
  if(serious_neck_chest_abdomen_injury(known_dominant_injury_pen[v,grepl( "AISCode" , names(known_dominant_injury_pen))]) >= 1 && known_dominant_injury_pen[v, "inj_dominant"] == 2) {
    known_dominant_injury_pen[v,"penetrating"] <- TRUE
  } else {
    known_dominant_injury_pen[v,"penetrating"] <- FALSE
  }
}
#penetrating <- TRUE/FALSE


#SEVERE TBI
head_region <- function(code) {
  region <- substr(as.character(code), 1, 1)
  is.element(region, 1)
}

#returns true if body region == 1

has_a_serious_head_injury <- function(code) {
  code <- code[!is.na(code)]
  serious_injury <- is_serious(code)
  head_region(code[serious_injury]) 
}

#acts on vectors in "serious_injury", chooses those which have region == 1

only_one_region <- function(code){
  number_of_regions(code) == 1
}

#generates TRUE/FALSE

only_one_serious_injury_region <- function(code) {
  code <- code[!is.na(code)]
  serious_injury <- is_serious(code)
  only_one_region(code[serious_injury])
}
#generates TRUE/FALSE

gcs_and_intub_is_na <- with(fewer_variables, is.na(ed_gcs_sum) & is.na(pre_intub_type))    
known_gcs_or_intub_type <- fewer_variables[!gcs_and_intub_is_na,]  
#remove row if na in both prehospital intub type and gcs

known_gcs_or_intub_type$gcs_below_9 <- NA

known_gcs_or_intub_type$gcs_below_9 <- with(known_gcs_or_intub_type, ifelse(ed_gcs_sum<=8 | pre_intub_type==1, TRUE, FALSE))
known_gcs_or_intub_type$gcs_below_9 <- with(known_gcs_or_intub_type, ifelse(is.na(gcs_below_9) | isFALSE(gcs_below_9), FALSE, TRUE))

#makes new variable containing those with a GCS <9 and those intubated in a prehospital setting. Converts na to false

known_gcs_or_intub_type$severe_tbi <- NA

for (i in 1:nrow(known_gcs_or_intub_type)) {
  v = 0+i
if (has_a_serious_head_injury(known_gcs_or_intub_type[v, grepl("AISCode", names(known_gcs_or_intub_type))]) == TRUE && only_one_serious_injury_region(known_gcs_or_intub_type[v, grepl("AISCode", names(known_gcs_or_intub_type))]) == TRUE && known_gcs_or_intub_type[v, "gcs_below_9"] == TRUE) {
    known_gcs_or_intub_type[v,"severe_tbi"] <- TRUE
  } else {
    known_gcs_or_intub_type[v,"severe_tbi"] <- FALSE
  }
}


#### MERGING 

blunt_multisystem_cohort <- filter(known_dominant_injury, blunt_multisystem == "TRUE")
shock_cohort <- filter(known_blood_pressure, shock == "TRUE")
penetrating_cohort <- filter(known_dominant_injury_pen, penetrating == "TRUE")
geriatric_cohort <- filter(known_age, geriatric == "TRUE")
severe_tbi_cohort <-filter(known_gcs_or_intub_type, severe_tbi == "TRUE")
overall <- fewer_variables
#creating cohort df:s

 geriatric_cohort$Cohort <- "Geriatric"
 penetrating_cohort$Cohort <- "Penetrating"
 severe_tbi_cohort$Cohort <- "Severe TBI"
 shock_cohort$Cohort <- "Shock"
 blunt_multisystem_cohort$Cohort <- "Blunt multisystem"
 overall$Cohort <- "Overall"
 #creating common variable "cohort"
 
  merged_cohorts <- Reduce(function(x, y) merge(x, y, all=TRUE),
                                     list(geriatric_cohort, penetrating_cohort, blunt_multisystem_cohort, shock_cohort, severe_tbi_cohort, overall))

  #merging cohorts

  
  
  
#### PLOTTING
  
  ## FLOWCHARTS                  
            
                  
  #SweTrau to M&M                
                  
 swetrau_flowchart <- DiagrammeR::grViz("
digraph graph2 {

graph [layout = dot]

# node definitions with substituted label text
node [shape = rectangle, width = 4, fillcolor = Biege]
a [label = '@@1']
b [label = '@@2']
c [label = '@@3']
d [label = '@@4']
e [label = '@@5']
f [label = '@@6']

a -> b -> c -> d -> e -> f 

}

[1]:  paste0('Trauma team activation following major trauma')
[2]: paste0('Inclusion in SweTrau')
[3]: paste0('Primary review and audit filters')
[4]: paste0('Secondary review by trained nurse')
[5]: paste0('Multidisciplinary mortality and morbidity conference')
[6]: paste0('Consensus regarding existence of OFI')
")                
  
#Stages of exclusion  
                        
   a <- c(0,1)               
   a$a <- nrow(problem_scrambled) 
   a <- as.data.frame(a)
   a$b <- nrow(known_problem_area)
   a$c <- nrow(fewer_variables)
   a$d <- nrow(blunt_multisystem_cohort)
   a$e <- nrow(penetrating_cohort)
   a$f <- nrow(severe_tbi_cohort)
   a$g <- nrow(shock_cohort)
   a$h <- nrow(geriatric_cohort)           
  #df with nrow of relevant df:s to be used as reference in code below 
   
exclusion_flowchart <- DiagrammeR::grViz("
digraph graph2 {

graph [layout = dot]

node [shape = rectangle, width = 4, fillcolor = Biege]
a [label = '@@1']
b [label = '@@2']
c [label = '@@3']
d [label = '@@4']
e [label = '@@4']
f [label = '@@5']
g [label = '@@6']
h [label = '@@7']


a -> b -> c -> d 
c -> e
c -> f
c -> g
c -> h
          
}

[1]: paste0('Trauma care quality database (n = ', a$a, ')')
[2]: paste0('Known problem area (n = ', a$b, ')')
[3]: paste0('Eligible for cohort inclusion (n = ', a$c, ')')
[4]: paste0('Blunt multisystem (n = ', a$d, ')')
[5]: paste0('Penetrating (n = ', a$e, ')')
[6]: paste0('Severe TBI (n = ', a$f, ')')
[7]: paste0('Shock (n = ', a$g, ')')
[8]: paste0('Geriatric (n = ', a$h, ')')

")

   
#### GROUPING AND EXCLUSION TABLE
   
Cohort <- c("Blunt multisystem", "Penetrating", "Severe TBI", "Shock", "Geriatric")
Eligible_for_inclusion <- c(nrow(blunt_multisystem_cohort), nrow(penetrating_cohort), nrow(severe_tbi_cohort), nrow(shock_cohort), nrow(geriatric_cohort))
Missing_data <- c((nrow(fewer_variables)-nrow(known_dominant_injury)),(nrow(fewer_variables)-nrow(known_dominant_injury)), (nrow(fewer_variables)-nrow(known_gcs_or_intub_type)), (nrow(fewer_variables)-nrow(known_blood_pressure)), (nrow(fewer_variables)-nrow(known_age)))
Type <- c("Dominant Injury", "Dominant Injury", "ED GCS", "ED blood pressure", "Age")
Exclusion_table <- data.frame(Cohort, Eligible_for_inclusion, Missing_data, Type)
   
Exclusion_table


   
 #### PATIENT DEMOGRAPHICS 
  
                  label(merged_cohorts$NISS)              <- "New Injury Severity Score"
                  label(merged_cohorts$pre_sbp_value)    <- "Pre Hospital Systolic BP"
                  label(merged_cohorts$ed_sbp_value)     <- "Emergency Department Systolic BP"
                  label(merged_cohorts$pt_age_yrs)        <- "Age"
                  
                  label(merged_cohorts$OFI)               <- "Opportunities for improvement"
                  
                  units(merged_cohorts$pre_sbp_value)  <- "mmHg"
                  units(merged_cohorts$ed_sbp_value)   <- "mmHg"
                  units(merged_cohorts$pt_age_yrs)    <-"Years"
                  

                  pt_demographics <- table1(~ OFI + pt_age_yrs + NISS + pre_sbp_value + ed_sbp_value | Cohort, data=merged_cohorts,overall = FALSE)
 #Since cohorts are merged in a manner allowing for duplicates, the "overall" number will not match the total number of participants, but instead the sum of the cohorts.  
                  
                  
                  
                  
                

#### INCIDENCE CALCULATIONS
   
                  
   fewer_variables$date <- as.Date(strptime(fewer_variables$Ankomst_te, format = "%Y%m%d %H:%M"))
#converting date               
                

### INCIDENCE (year)

cohorts <- list(geriatric_cohort = geriatric_cohort,
                 penetrating_cohort = penetrating_cohort,
                 blunt_multisystem_cohort = blunt_multisystem_cohort,
                 shock_cohort = shock_cohort,
                 severe_tbi_cohort = severe_tbi_cohort,
                 overall = overall)
 calculate_incidence <- function(cohort) {
     cohort$date <- as.Date(strptime(cohort$Ankomst_te, format = "%Y%m%d %H:%M"))
     cohort.ofi <- incidence(cohort$date, interval = "year", groups = cohort$OFI) %>% as.data.frame()
     cohort.ofi$incidence <- 0
     if (is.element("NO", names(cohort.ofi)))
         cohort.ofi$incidence <- cohort.ofi$YES / (cohort.ofi$YES + cohort.ofi$NO) * 100
     return (cohort.ofi)
 }
 incidence.data <- lapply(cohorts, calculate_incidence)


### CUMULATIVE INCIDENCE (year)

 calculate_percent_of_total <- function(cohort) {
     cohort$date <- as.Date(strptime(cohort$Ankomst_te, format = "%Y%m%d %H:%M"))
     cohort.ofi <- incidence(cohort$date, interval = "year", groups = cohort$OFI) %>% as.data.frame()
     cohort.ofi$per_tot <- 0
     if (is.element("NO", names(cohort.ofi)))
         cohort.ofi$per_tot <- cohort.ofi$YES / c(nrow(cohort)) * 100
     return (cohort.ofi)
 }
     
  percentage.data <- lapply(cohorts, calculate_percent_of_total)    
 
 get_cum_inc<- function(cohort){
cum_inc <- transform(cohort, cum_inc = cumsum(cohort$per_tot))
 return(cum_inc)
 }
 
cumulative.incidence.data <- lapply(percentage.data, get_cum_inc)
 




#### RENAMING VARIABLES

incidence_data <- incidence.data
incidence.data$geriatric_cohort$incidence_ger <- incidence.data$geriatric_cohort$incidence
incidence.data$penetrating_cohort$incidence_pen <- incidence.data$penetrating_cohort$incidence
incidence.data$severe_tbi_cohort$incidence_tbi <- incidence.data$severe_tbi_cohort$incidence
incidence.data$shock_cohort$incidence_sho <- incidence.data$shock_cohort$incidence
incidence.data$blunt_multisystem_cohort$incidence_blu <- incidence.data$blunt_multisystem_cohort$incidence
incidence.data$overall$incidence_all <- incidence.data$overall$incidence

#making unique variable name for each cohort (there surely is a way to do this more efficiently, but right now it is not my focus

cumulative.incidence.data$geriatric_cohort$cum_inc_ger <- cumulative.incidence.data$geriatric_cohort$cum_inc
cumulative.incidence.data$penetrating_cohort$cum_inc_pen <- cumulative.incidence.data$penetrating_cohort$cum_inc
cumulative.incidence.data$severe_tbi_cohort$cum_inc_tbi <- cumulative.incidence.data$severe_tbi_cohort$cum_inc
cumulative.incidence.data$shock_cohort$cum_inc_sho <- cumulative.incidence.data$shock_cohort$cum_inc
cumulative.incidence.data$blunt_multisystem_cohort$cum_inc_blu <- cumulative.incidence.data$blunt_multisystem_cohort$cum_inc
cumulative.incidence.data$overall$cum_inc_all <- cumulative.incidence.data$overall$cum_inc

#no2


cumulative.incidence.data <- bind_rows(cumulative.incidence.data, .id = NULL)
cumulative.incidence.data[is.na(cumulative.incidence.data)] = 0
cumulative.incidence.data<- ddply(cumulative.incidence.data,"dates",numcolwise(sum))
#list to df

incidence.data <- bind_rows(incidence.data, .id = NULL)
incidence.data[is.na(incidence.data)] = 0
incidence.data<- ddply(incidence.data,"dates",numcolwise(sum))
#list to df




#### PLOTTING INCIDENCE AND CUMULATIVE INCIDENCE

#Yearly incidence as %
plot_inc_year <- incidence.data %>% select(dates, incidence_ger, incidence_pen, incidence_blu, incidence_sho, incidence_tbi, incidence_all)

plot_inc_year_long <- melt(plot_inc_year, id.vars = "dates")
#melt as to enable ggplot to work

incidence_plot <- ggplot(plot_inc_year_long,
       aes(x=dates,
           y=value,
           col = variable)) + 
  geom_line(size = 1.0, alpha = 1.0) +
  labs(title = "Yearly incidence of OFI in the KUH trauma care quality database",
       x = "Year",
       y = "%") +
  theme_minimal() +
  theme(axis.title = element_text())
#Right now not showing S-TBI cohort as there are missing values.


#Cumulative incidence
plot_cum_inc<-cumulative.incidence.data %>% select(dates, cum_inc_ger, cum_inc_pen, cum_inc_blu, cum_inc_sho, cum_inc_tbi, cum_inc_all)

plot_cum_inc_long <- melt(plot_cum_inc, id.vars = "dates")

cum_incidence_plot<- ggplot(plot_cum_inc_long,
       aes(x=dates,
           y=value,
           col = variable)) + 
  geom_line(size = 1.0, alpha = 1.0) +
  labs(title = "Cumulative incidence of OFI in the KUH trauma care quality database by year",
       x = "Year",
       y = "% of total") +
  theme_minimal() +
  theme(axis.title = element_text())



#### INCIDENCE TRENDS

library(fable)

calculate_incidence_trends <- function(cohort) {
   trend <- as_tsibble(cohort) %>% model(lm = TSLM(incidence ~ trend() + season())) %>% tidy()
     return (trend)
 }

incidence.trends <- lapply(incidence_data, calculate_incidence_trends)


penetrating_incidence_trend <- incidence.trends$penetrating_cohort[c(2), ] 
geriatric_incidence_trend <- incidence.trends$geriatric_cohort[c(2), ] 
tbi_incidence_trend <- incidence.trends$severe_tbi_cohort[c(2), ] 
blunt_incidence_trend <- incidence.trends$blunt_multisystem_cohort[c(2), ] 
shock_incidence_trend <- incidence.trends$shock_cohort[c(2), ] 
overall_incidence_trend <- incidence.trends$overall[c(2), ] 


penetrating_incidence_trend$cohort <- "Penetrating"
geriatric_incidence_trend$cohort <-"Geriatric"
tbi_incidence_trend$cohort <- "Severe TBI"
blunt_incidence_trend$cohort <- "Blunt multisystem"
shock_incidence_trend$cohort <- "Shock"
overall_incidence_trend$cohort <- "Overall"

incidence_trend_table <- Reduce(function(x, y) merge(x, y, all=TRUE),
                                     list(penetrating_incidence_trend, geriatric_incidence_trend, tbi_incidence_trend, blunt_incidence_trend, shock_incidence_trend, overall_incidence_trend))

library(dplyr)
library(readxl)

names(incidence_trend_table) <- c('Model', 'Term', 'Estimate', 'Cohort', 'SD_error', 'Statistic', 
                       'P_value')
incidence_trend_table <- incidence_trend_table[,c(4, 3, 5, 6, 7, 2, 1)]
incidence_trend_table <- incidence_trend_table[, -c(6,7)]

### GRAPH PRESENTING YEARLY M&M PARTICIPANTS AS % OF YEARLY SWETRAU PARTICIPANTS 

swetrau_scrambled_date <- swetrau_scrambled
swetrau_scrambled_date$date <-substr(as.character(swetrau_scrambled$DateTime_ArrivalAtHospital),1,10)
swetrau_scrambled_date$date <- as.Date(swetrau_scrambled_date$date)

swetrau_scrambled_date <- swetrau_scrambled_date %>% select(id, date)

swetrau_scrambled_date <- incidence(swetrau_scrambled_date$date, interval = "year")


problem_scrambled_date <- problem_scrambled
problem_scrambled_date$date <-as.Date(strptime(problem_scrambled_date$Ankomst_te, format = "%Y%m%d %H:%M"))
#converting date

problem_scrambled_date <- problem_scrambled_date %>% select(id, date)

problem_scrambled_date <- incidence(problem_scrambled_date$date, interval = "year")
do.call(rbind.data.frame, problem_scrambled_date)
do.call(rbind.data.frame, swetrau_scrambled_date)

problem_scrambled_date <- as.data.frame(problem_scrambled_date)
swetrau_scrambled_date <- as.data.frame(swetrau_scrambled_date)

problem_swetrau_percentage <- merge(problem_scrambled_date, swetrau_scrambled_date, by = "dates")

problem_swetrau_percentage$MM_percentage_of_swetrau <- problem_swetrau_percentage$counts.x/ problem_swetrau_percentage$counts.y *100

ggplot(problem_swetrau_percentage,
       aes(x=dates,
           y=MM_percentage_of_swetrau)) + 
  geom_line(size = 1.0, alpha = 1.0) +
  labs(title = "M&M as percent of SweTrau",
       x = "Year",
       y = "% of total") +
  theme_minimal() +
  theme(axis.title = element_text())

   
### INCIDENCE PACKAGE PLOT 

fewer_variables$date <- as.Date(strptime(fewer_variables$Ankomst_te, format = "%Y%m%d %H:%M"))
#converting date

incidence_package_ofi <- incidence(fewer_variables$date, interval = "year", groups = fewer_variables$OFI)
#Use of incidence package to calculate incidence in the group as a whole by year (2)

plot(incidence_package_ofi, stack = TRUE, border = "grey")
### TOTAL

#calculate incidence by cohort and year (2)
cohort_incidence_package_ofi <- filter(merged_cohorts, OFI == "YES")

cohort_incidence_package_ofi$date <-as.Date(strptime(cohort_incidence_package_ofi$Ankomst_te, format = "%Y%m%d %H:%M"))
#converting date

#Since each case here is "OFI == YES", incidence of dates is all we need to calculate

cohort_incidence_package_ofi <-incidence(cohort_incidence_package_ofi$date, interval = "year", groups = cohort_incidence_package_ofi$Cohort)

plot(cohort_incidence_package_ofi)
### BY COHORT

#### References for markdown code

cum_inc <- select(plot_cum_inc, -dates)
cum_inc <- cum_inc %>% slice(-c(1:8))


incidence <- select(incidence.data, incidence_ger, incidence_pen, incidence_blu, incidence_sho, incidence_tbi)

min.col <- function(m, ...) max.col(-m, ...)


```


## Results
### Participants 
Over the pre-specified study period, `r nrow(problem_scrambled)` patients were reviewed at a multidisciplinary mortality and morbidity conference at KUH and included in the trauma care quality database. Of those, `r nrow(fewer_variables)` (`r nrow(fewer_variables)/nrow(problem_scrambled)*100`%) patients had sufficient data for grouping into cohorts and went on to form the study population (see Figure 1). Participants were then divided into the five cohorts blunt multisystem (n=`r nrow(blunt_multisystem_cohort)`), geriatric (n=`r nrow(geriatric_cohort)`), penetrating (n= `r nrow(penetrating_cohort)`), severe TBI (n=`r nrow(severe_tbi_cohort)`) and shock (n=`r  nrow(shock_cohort)`). `r Exclusion_table$Missing_data[1]` patients lacked data about dominant injury type; `r Exclusion_table$Missing_data[3]` patients lacked data about emergency department gcs; `r Exclusion_table$Missing_data[4]` patients lacked data about emergency department blood pressure. No patients lacked data about age (see Table 1).


`r exclusion_flowchart`
**Figure.1. Exclusion flowchart**




`r kable(Exclusion_table)`

**Table.1. Missing data by cohort**
As a substantial number of patients lack data on one or several data points, missing data was not used to categorically exclude patients. In table 1, the variable "Missing data" refers to the number of patients lacking sufficient data for grouping into a specified cohort.
 
 
### Descriptive data
Table 2 presents patient demographics in the trauma cohorts. The cohorts exhibited differences regarding age, injury severity score, emergency department systolic BP, (insert extra variables). The cohort with the highest percentage of OFI was `r colnames(cum_inc)[apply(cum_inc,1,which.max)]` with `r max(cum_inc)`%. The cohort with the lowest percentage of OFI was `r colnames(cum_inc)[apply(cum_inc,1,which.min)]` with `r min(cum_inc)`%.

`r pt_demographics`

**Table.2. Patient demographics by cohort**


### Incidence
In the study population, total incidence of OFI ranged from  `r min(incidence.data$incidence_all)`% to   `r max(incidence.data$incidence_all)`% (see Figure 2). The highest recorded yearly incidence was `r max(incidence)`% in the `r colnames(incidence)[apply(cum_inc,1,which.max)]` cohort. Overall, there was a narrow/wide distribution of incidence of OFI across cohorts. The cumulative incidence rate varied significantly between cohorts. The cohort with the largest cumulative incidence was `r colnames(cum_inc)[apply(cum_inc,1,which.max)]` (`r max(cum_inc)`%)  and the cohort with the lowest was `r colnames(cum_inc)[apply(cum_inc,1,which.min)]` (`r min(cum_inc)`%). The cumulative incidence in the population as a whole was`r max(cum_inc$cum_inc_all)`%.

```{r echo=FALSE, warning = FALSE}
library(ggplot2)
incidence_plot
```

**Figure.2. Yearly incidence of OFI**
Percent of yearly cohort participants with OFI.


```{r echo=FALSE, warning = FALSE}
library(ggplot2)
cum_incidence_plot
```

**Figure.3. Cumulative incidence of OFI**
 Cumulative percent of total cohort participants with OFI each year.


### Trends
The cohort with the most significant downtrend was XXX with a p-value of XXX...

`r kable(incidence_trend_table)`

**Table.3. Incidence trends **
